EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-11-08 BrainScience/brainscience.tex
Toques

A problem with any P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.  Therefore, P300 experiments are often mined with \textit{null-signals}.  These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the real signal is not generated.   This lack of certainty may be in detriment of analysis of algorithms and can be misleading or difficult to deal with.   Similar experiments had addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.

To solve this issue a pseudo-real dataset is generated under two different modalities.  A passive modality and an active modality.

@@

First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into a distinct EEG stream.  A set of template ERPs are acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. On the other hand, the EEG traces where these templates are superimposed, are experimentally obtained by subjects which are observing the flashing of the stimulus matrix during a P300 Speller procedure but they do not engage in focusing on any letter in particular. Everything is there, except the P300 ERP component. Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.  By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential.

A sample P300 ERP obtained from the trial number $2$ of Subject 8 can be seen in Figure  \ref{fig:erptemplate1}.  The templates obtained by this processing this Subject's EEG signals are selected due to their shapes more closely resembling the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.  Segments are extracted from the original dataset, and they are point-to-point coherently averaged.

After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha

@@

Second, an active-modality is also used, where a P300-Based BCI experiment is performed on four subjects, but the signal segments are modified to produce the desired P300 response, but based on information extracted from the same subject.
In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.  The most prototypical P300 waveform template is selected by visual observation.  This template is superimposed to each segment marked as \textit{True} on the EEG stream.  The baseline of each segment, on the other hand, is obtained by picking randomly any \textit{False} segment for the same subject.

@@

\subsubsection{EEG Stream Generation}

An EEG stream with null-P300 signal is obtained by the following procedure:

The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.  Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.  This yields exactly $120$ intensifications of rows and columns per trial.  The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.  This provides a $4$ Hz frequency of flashes on the screen.  The initial pause and the inter-trial pauses are set to $20$ seconds.  The whole experiment lasts for around $1400$ seconds.

The remaining four participants perform a copy-spelling task where the computer monitor suggests by highlighting the target letter, which letter the subject needs to focus.  During each selection letter trial, the target letter is suggested at the bottom of the screen.

>>>>

A problem with any P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.  Therefore, P300 experiments are often mined with \textit{null-signals}.  These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the expected signal component is not generated.   This lack of certainty may be in detriment of any conducted analysis and can be misleading or difficult to deal with.   Previous works have addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.

In order to tackle this problem, a pseudo-real dataset based on an EEG stream is generated under two different modalities.  A passive modality and an active modality.

\subsubsection{EEG Stream Generation}
\label{EEGStream}

The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.  Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.  This yields exactly $120$ intensifications of rows and columns per trial.  The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.  This provides a $4$ Hz frequency of flashes on the screen.  The initial pause and the inter-trial pauses are set to $20$ seconds.  The whole experiment lasts for around $1400$ seconds.  This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as True and the remaining $700$ as False.

The remaining four participants perform a copy-spelling task where the computer monitor highlights the target letter, which is the one that the subject needs to focus.  Across the duration of the trial, the current target letter is informed at the bottom of the screen.

@@

First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into a distinct EEG stream.  A set of template ERPs are acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. The experimental protocol implemented to produce this dataset is the same as the one described in Section~\ref{EEGStream}. On the other hand, the EEG traces where these templates are superimposed, are experimentally obtained by subjects which are observing the flashing of the stimulus matrix during a P300 Speller procedure but they do not engage in focusing on any letter in particular. Everything is there, except the P300 ERP component. Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.  By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential.

A sample P300 ERP obtained from the trial number $2$ of Subject 8 can be seen in Figure  \ref{fig:erptemplate1}.  These templates are selected due to their shapes more closely resembling the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.  They are produced by extracting segments for this subject and by point-to-point coherently average them.

%After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha

@@

Second, an active modality is also implemented, where a P300-Based BCI Speller experiment is performed on four subjects. For this scenario, the signal segments are modified to guarantee the inclusion of a P300 component. However, in this case the templates are extracted from the same subject.  Hence, the EEG signal is preprocessed and labeled segments are extracted.  Segments labeled \textit{True} are coherently point-to-point averaged, and $70$ templates are produced from the whole set of $35$ trials.

Once templates are procured, a random \textit{False} segment for the same subject is obtained.  This is used as a baseline signal and is added to the template, conforming a new segment which contains one of the templates.  This procedure continues until the $700$ segments marked as \textit{True} are completed.

%The EEG stream contains $4200$ segments, which correspond to the preprocessed EEG data that was obtained coincidentally with the intensification of a row or column on the Speller Matrix.

@@
